# An application to Vietnamese climate data {#sec-appliout}

## Data description and preprocessing {#sec-datades}

```{r fig-vnt_climate_regions_map}
#| fig-cap: The three climate regions of Northern Vietnam.
#| fig-width: 11.25
#| fig-height: 7.5

library(dda)
library(sf)
library(tidyverse)

vietnam_provinces <- vietnam_provinces |>
  arrange(region, province) |>
  mutate(climate_region = case_when(
    province %in% c("LAICHAU", "DIENBIEN", "SONLA") ~ "S1",
    province %in% c(
      "LAOCAI", "YENBAI", "HAGIANG", "CAOBANG", "TUYENQUANG",
      "BACKAN", "THAINGUYEN", "LANGSON", "BACGIANG", "QUANGNINH"
    ) ~ "S2",
    region %in% c("NMM", "RRD") ~ "S3",
    .default = NA
  ))

vietnam_provinces |>
  filter(region %in% c("RRD", "NMM")) |>
  ggplot(aes(fill = climate_region, label = province)) +
  geom_sf() +
  geom_sf_label(fill = "white") +
  scale_fill_viridis_d() +
  labs(x = "longitude", y = "latitude", fill = "climate region")
```

In this application, we study daily maximum temperatures for each of the
$I=63$ Vietnamese provinces over a $T=30$-year period (1987-2016).
Originally from the Climate Prediction Center (CPC) database, developed
and maintained by the National Oceanic and Atmospheric Administration
(NOAA), the data underwent a preliminary treatment presented in
[@trinh_discrete_2023]. From the daily 365 or 366 values for each year,
we derive the yearly maximum temperature distribution for each of the
1,890 province-year units. We assume that the temperature samples are
independent across years and spatially across provinces, which is a
simplifying assumption. @fig-vnt_north_south_map depicts the six administrative
regions of Vietnam, and the corresponding provinces. However, these
regions cover areas with varied climates. To achieve more climatically
homogeneous groupings, we use clusters of provinces based on climatic
regions as defined by @stojanovic_trends_2020. @fig-vnt_climate_regions_map displays the three climatic regions
covering Northern Vietnam. We focus on region S3, composed of 13
provinces, by similarity with the North Plain (Red River Delta) (S3) in
[@stojanovic_trends_2020].

@fig-vnt_climate_regions_dens shows the maximum temperature densities for the 13 provinces of S3,
plotted by year, using the preprocessing detailed in @sec-preproc with degree less than or equal to $d=4$,
smoothing parameter $\lambda=10$ and $k=10$ knots located at quantiles
of the pooled sample (across space and time). We observe more
variability across time than across space which confirms that the
spatial homogeneity objective is achieved.

```{r fig-vnt_climate_regions_dens}
#| fig-cap: Maximum temperature densities for the 13 provinces in the S3 climate
#|   region of Northern Vietnam, 1987-2016, colour-coded by province.

vntcr <- vietnam_temperature |>
  filter(region %in% c("RRD", "NMM")) |>
  dplyr::select(year, province, t_max) |>
  left_join(vietnam_provinces, by = "province") |>
  st_as_sf() |>
  dplyr::select(year, climate_region, province, t_max) |>
  filter(climate_region == "S3") |>
  arrange(year, province)

vnt <- vntcr |>
  mutate(t_max = as_dd(t_max, lambda = 10, nbasis = 10, mc.cores = 15))
class(vnt$t_max) <- c("ddl", "fdl", "list")

vnt |> plot_funs(t_max, color = province) +
  facet_wrap(vars(year)) +
  labs(x = "temperature (deg. Celsius)", y = "density")
```

## Outlier detection using ICS for the S3 climate region of Vietnam

We follow the different steps described in @sec-icsout,
and examine the results of ICS outlier detection using the scatter pair
$(\operatorname{Cov}, \operatorname{Cov}_4)$ on the 390 (13 provinces
$\times$ 30 years) densities from region S3, obtained after the
preprocessing detailed above.

```{r fig-vnt_climate_regions_screedist}
#| fig-cap: Scree plot of the ICS eigenvalues (left panel), and the ICS
#|   distances based on the first two components (right panel) for maximum
#|   temperature densities for the 13 provinces in the S3 climate region of
#|   Northern Vietnam, 1987-2016.
#| fig-subcap: ""
#| fig-width: 7.5
#| fig-height: 5
#| layout-ncol: 2

icsout <- ICS_outlier(vnt$t_max, index = 1:2, n_cores = 14)

icsout$result |> ggplot(aes(index, gen_kurtosis, group = 1)) +
  geom_line(alpha = 0.5) +
  geom_point(aes(color = index, alpha = selected), size = 3) +
  theme_legend_inside_right +
  labs(x = "", y = "ICS eigenvalues") +
  guides(color = "none")

vnt <- vnt |>
  mutate(
    index = seq_len(nrow(icsout$scores)),
    outlying = as.factor(icsout$outliers),
    distances = icsout$ics_distances
  ) |>
  cbind(icsout$scores)

vnt |>
  mutate(province = ifelse(outlying == 1, province, NA)) |>
  ggplot(aes(year, distances, color = province, label = year)) +
  geom_text() +
  scale_color_discrete(limits = function(x) ifelse(is.na(x), "", x)) +
  geom_hline(yintercept = icsout$ics_dist_cutoff) +
  labs(color = "outlying\nprovinces")
```

```{r fig-vnt_climate_regions_eigdensscat}
#| fig-cap: Scatter plot of the first two invariant components (left
#|   panel) labelled by year and coloured by province, and the first two ICS
#|   dual eigendensities (right panel) of the maximum temperature densities
#|   for the 13 provinces in the S3 climate region of Northern Vietnam,
#|   1987-2016.
#| fig-subcap: ""
#| fig-width: 7.5
#| fig-height: 5
#| layout-ncol: 2

vnt |>
  mutate(province = ifelse(outlying == 1, province, NA)) |>
  ggplot(aes(IC.1, IC.2)) +
  scale_color_discrete(limits = function(x) ifelse(is.na(x), "", x)) +
  geom_text(aes(color = province, label = year)) +
  labs(color = "outlying\nprovinces")

icsout$result |>
  filter(selected) |>
  plot_funs(H_dual, color = index) +
  geom_hline(yintercept = 1 / diff(icsout$result$H_dual[[1]]$basis$rangeval), linetype = "dotted") +
  labs(x = "temperature (deg. Celsius)", y = "density") +
  scale_color_manual(values = scales::pal_hue()(9)[1:2]) +
  theme_legend_inside
```

The scree plot on the left panel of @fig-vnt_climate_regions_screedist clearly indicates that we
should retain the first two invariant components. The right panel of
@fig-vnt_climate_regions_screedist shows the squared ICS
distances based on these first two components, with the observations
index on the $x$-axis and with a threshold (horizontal line)
corresponding to a significance level of $2.5\%$. This plot reveals that
several observations are distinctly above this threshold, especially for
the years 1987 and 2010.

The left panel of @fig-vnt_climate_regions_eigdensscat displays the scatter plot
of the first two components, labelled by year. The densities are
coloured by province for the outliers and coloured in grey for the other
provinces. This plot reveals that the outliers are either densities from
2010 (and one density from 1998) that are outlying on the first
component, or densities from 1987 and 2007 that are outlying on the
second component.

To interpret the outlyingness, we can use the dual eigendensities
plotted in the right panel of @fig-vnt_climate_regions_eigdensscat together with @fig-vnt_climate_regions_densout, which represents the densities
and their centred log-ratio transformation, colour-coded by year for the
outliers and in grey for the other observations. This is justified by
the reconstruction formula of @prp-reconstruction in the Appendix. The horizontal line on
the eigendensities plot (right plot of @fig-vnt_climate_regions_eigdensscat) corresponds to the uniform
density on the interval $[5;40]$. Four provinces in 2010 are outlying
with large positive values on the first invariant component (see the
left panel of @fig-vnt_climate_regions_eigdensscat). The first eigendensity
IC.1 is characterised by a smaller mass of the temperature values on the
interval $[5;20]$, compared to the uniform distribution, a mass similar
to the uniform on $[20;35]$, and a much larger mass than the uniform on
the interval $[35;40]$. These four observations correspond to the four
blue curves on the left and right panels of @fig-vnt_climate_regions_densout. Compared to the other
densities, these four densities exhibit relatively lighter tails on the
lower end of the temperature spectrum and heavier tails on the higher
end. For temperature values in the medium range, these four observations
fall in the middle of the cloud of densities and of clr transformed
densities. On the second invariant component, six observations take
large values and are detected as outliers. They correspond to four
provinces in 1987 and three in 2007 (see the left panel of @fig-vnt_climate_regions_eigdensscat). The second eigendensity
IC.2 differs greatly from the uniform distribution on the whole interval
of temperature values. The left tail is much lighter while the right
tail is much heavier. Besides the six observations flagged as outliers,
other provinces in 1987 and 2007 take large values on IC.2, and
correspond to densities with very few days with maximum temperature less
than 15 degrees Celsius compared to other densities.

```{r fig-vnt_climate_regions_densout}
#| fig-cap: Maximum temperature densities (left panel) and their centred
#|   log-ratio transforms (right panel) for the 13 provinces in the S3
#|   climate region of Northern Vietnam, 1987-2016, outlying densities are
#|   colour-coded by year.
#| fig-subcap: ""
#| fig-width: 7.5
#| fig-height: 5
#| layout-ncol: 2

vnt |>
  mutate(year = as.factor(ifelse(outlying == 1, year, NA))) |>
  arrange(outlying) |>
  plot_funs(t_max, color = year, alpha = outlying) +
  theme_legend_inside +
  labs(
    x = "temperature (deg. Celsius)", y = "density",
    color = "outlying\nyears"
  ) +
  scale_color_discrete(limits = function(x) ifelse(is.na(x), "", x)) +
  guides(alpha = "none")

vnt |>
  mutate(
    t_max = as.fd(center(t_max)),
    year = as.factor(ifelse(outlying == 1, year, NA))
  ) |>
  arrange(outlying) |>
  plot_funs(t_max, color = year, alpha = outlying) +
  theme(legend.position = "none") +
  labs(
    x = "temperature (deg. Celsius)", y = "centred clr(density)",
    color = "outlying\nyears"
  )
```

## Influence of the preprocessing parameters {#sec-app_smooth}

```{r fig-vnt_climate_regions_grid_outliers}
#| fig-cap: Outlier detection by ICS across smoothing parameters for the Vietnam
#|   climate data. *Top:* 2 invariant components selected; *Bottom:* automatic
#|   selection through D'Agostino tests. *$y$-axis:* year; *$x$-axis:* $\lambda$
#|   parameter. Columns correspond to knot numbers (5-25). Outliers are
#|   marked as light gray to black squares depending on their detection
#|   frequency.
#| fig-width: 11.25
#| fig-height: 7.5

knots_pos_seq <- "quantiles"
nknots_seq <- 5 * 1:5
lambda_seq <- 10^(-2:2)
index_seq <- c(2, NA)

param_grid <- expand.grid(
  knots_pos = knots_pos_seq,
  nknots = nknots_seq,
  lambda = lambda_seq,
  index = index_seq
) |>
  arrange(nknots, knots_pos, lambda)

if (!file.exists("data/vnt_climate_regions.RData")) {
  system.time(
    ics_grid <- param_grid |>
      mutate(id = seq_len(n())) |>
      rowwise() |>
      mutate(res = list(smooth_ICS(vntcr, knots_pos, nknots, lambda, index)))
  )
  save(ics_grid, file = "data/vnt_climate_regions.RData")
}

load("data/vnt_climate_regions.RData")

ics_grid_out <- ics_grid |>
  rowwise() |>
  filter(!(any(class(res$ics_out) == "error"))) |>
  mutate(res = list(data.frame(
    obs_id = seq_len(length((res$ics_out$outliers))),
    outlying = res$ics_out$outliers == 1
  ))) |>
  unnest(res)

ics_grid_out |>
  mutate(index = ifelse(is.na(index), "auto. selection of ICs",
    paste(index, "ICs selected")
  )) |>
  rename(index_select = index) |>
  left_join(vnt |> mutate(obs_id = row_number()) |> select(!outlying), by = c("obs_id")) |>
  group_by(nknots, lambda, index_select, year) |>
  mutate(n_outlying = sum(outlying) / 13) |>
  ggplot(aes(log10(lambda), year, fill = n_outlying, alpha = n_outlying)) +
  geom_tile() +
  scale_alpha_continuous(range = c(0, 1)) +
  scale_fill_gradient(low = "white", high = "black") +
  facet_grid(cols = vars(nknots), rows = vars(index_select)) +
  labs(y = "year", fill = "outlier\nfrequency", alpha = "") +
  guides(alpha = "none")
```

```{r fig-vnt_climate_regions_grid_summary}
#| fig-cap: Frequency of outlier detection by ICS across all 25 scenarios with
#|   varying smoothing parameters and all 13 provinces, for each year in the
#|   Vietnamese climate dataset.
#| fig-width: 11.25
#| fig-height: 7.5

ics_grid_summary <- ics_grid_out |>
  mutate(outlying = ifelse(outlying, 1, 0)) |>
  group_by(obs_id) |>
  summarise(n_outlying = sum(outlying)) |>
  left_join(vnt |> mutate(obs_id = row_number()), by = c("obs_id")) |>
  select(obs_id, province, year, n_outlying)

ics_grid_summary |>
  ggplot(aes(year, n_outlying)) +
  geom_bar(stat = "identity") +
  labs(y = "outlier detection frequency")
```

As mentioned in @sec-toyex, we can validate the atypical nature of
observations by running the ICSOutlier procedure multiple times with
varying smoothing parameter values. Following the rule of thumb of one
dimension per 10 observations, with 390 observations, we should consider
less than 35 interior knots. In what follows, we take 5, 10, 15, 20 and
25 interior knots and we consider base-10 logarithm values for $\lambda$
equal to -2, -1, 0, 1 and 2. The number of selected ICS components is
either fixed equal to 2, or is automatically determined using the
D'Agostino normality test described in @sec-icsout.
We compute the squared ICS distances of the 390 observations, and
observations are classified as outliers when their squared distance
exceeds the threshold based on a $2.5\%$ level as detailed in @sec-icsout.

We plot in @fig-vnt_climate_regions_grid_outliers the years on the $y$-axes for
the 25 smoothing parameter setups, indicating outlying years with light
gray to black squares depending on their detection frequency. @fig-vnt_climate_regions_grid_summary displays a bar plot of
the frequency of outlier detection (across the 25 setups and the 13
provinces) for each year. Note that the choice of the number of selected
invariant components has minimal impact. Both @fig-vnt_climate_regions_grid_outliers and
@fig-vnt_climate_regions_grid_summary confirm the results of
the previous section. Most provinces are outlying in 1987 and several
are also outlying in 2007 and 2010. For large values of $\lambda$, many
provinces are also detected as outliers in 2016. Some provinces are
detected quite often over the years: THANH HOA, HAI PHONG and HOA BINH.
Note that in [@stojanovic_trends_2020], the province of THANH HOA
extends across two climatic regions (S3 and S4) which could explain why
it is very often detected as an outlier.

An overall comment regarding the outlier detection procedure that we use
in the present application is that, from our experience on other data
sets, an outlying density is often characterised by a behaviour that
differs from the other densities in the tails of the distribution. This
is not surprising because the Bayes inner product defined by equation
@eq-innerpdt involves the ratio of densities which can be
large when a density is small (at the tails of the distribution).
